---
title: "Gaming_Study_DSasField_Final_Project"
author: "Lauren Bain, Shyam Vangeepuram, Anna Vernier"
date: "2/9/2022"
output: pdf_document
---
## Project Purpose:

This analysis explores the potential mental health implications of time spent online gaming. Additionally, we explore the types of people who are playing, and what games they play.

## Introduction:

In the last few decades, playing video games online has dramatically increased in popularity. Correlations have been observed between excessive or compulsive gaming behavior and psychological disorders such as depression, anxiety, social phobia, and decreased life satisfaction, especially among adolescents. However, most of the existing research literature into the connections between mental health and excessive video gaming have used small sample sizes and focused primarily on young adults from limited cultural backgrounds.

The dataset used for this analysis consists of responses from over 13,000 participants in an international survey. Participants were solicited in the subreddit /r/league of legends, a popular multiplayer game in which teams of ten players battle. Predictably, This game comprised the largest sub-sample of players in the dataset.
The survey evaluated gaming behavior across several dimensions, including what game respondents played most, their preferred platform, and whether they played alone, with strangers, or with friends. Subjects were also asked to estimate how many hours they spent playing in an average week, and how many hours they spent engaged with gaming outside of play, participating in activities such as watching streams, replays, or tutorials.

Participants also answered questions from screening assessments of general anxiety disorders, social phobia, overall life satisfaction, and narcissism.

### Anxiety

Anxiety was measured using the Generalized Anxiety Disorder 7 scale (GAD-7; Spitzer et al, 2006) which measures how severely the person has experienced seven symptoms on a scale of 0-5 (nervousness, inability to stop worrying, excessive worry, restlessness, difficulty in relaxing, easy irritation, and fear of something awful happening) in the last two weeks. The points are summed for a normative total score, with 5, 10, and 15 representing mild, moderate, and severe anxiety respectively.

### Life Satisfaction

The Satisfaction with Life Scale (SWLS; Diener et al, 1985) is a five-item screening with a seven-point scale ranging from “strongly disagree” to “strongly agree” on statements like, “I am satisfied with life.” It is designed to interrogate individuals’ subjective experience of their quality of life without specifying domains such as health or employment status. Each item is summed for a score of 5-35.

### Social Phobia

Although the GAD-7 test also broadly addresses social anxiety, the Social Phobia InveNtory (SPIN; Antony et al, 2006) was included in this survey to specifically assess fear, avoidance, and physiological arousal in the preceding week. This screening has 17 items such as “I avoid going to parties” and “talking to strangers scares me.” Respondents are asked to evaluate how true each item is for them from 0 “not at all” to 4 “extremely” and is scored from 0-68, with below 20 interpreted as no social anxiety, 21-30 as mild, 31-40 as moderate, and 41-50 as severe, and 51+ as very severe.

### Narcissism

Finally, the Single Item Narcissism Scale (SINS; Konrath et. Al, 2014) simply asked participants, “On a scale of 0-7, how true is this statement: 'I am a narcissist.’ (note that ‘narcissist’ means egotistical, self-focused, and vain).” Although not adequate for clinical diagnosis, this test has been validated as reasonably accurate against more extensive screenings for narcissism, and is considered an effective measure in studies where time or other constraints make more extensive screenings less feasible. The inclusion of the SINS measure in this survey is interesting because narcissism has been linked to both lower anxiety, and difficulty maintaining interpersonal relationships. As such, this item acts as a mitigating factor between the measures of social phobia and general anxiety.

## Data Details

In the data set, each answer to the psychological screening questionnaires were kept separate. Before we began our analysis, we summed the results of each test following the recommended guideline criteria into single variables so that each subject had a total GAD-7, SWLS, and SPIN score. Where applicable, we placed the scores in bins to narrow them down for simpler
analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Packages used
library(tidyverse)
library(readr)
library(ggplot2)
library(gridExtra)
library(maps)
library(rworldmap)
library(viridis)


#Reading in data:
#Note: csv file needs to be downloaded and imported by user.

GamingStudy_data <- read_csv("GamingStudy_data.csv")

#data frame to be used
gsd <- GamingStudy_data
```

## Data Cleaning

```{r}
#remove excess columns
gsd <- GamingStudy_data %>% select(-c('S. No.', 'Timestamp'))

#remove outliers
gsd <- gsd[!(gsd$Hours > 120),]
gsd <- gsd[!(gsd$streams > 120),]

###GAD Scores

#sum GAD scores
gsd$Total_GAD_Score <- rowSums(gsd[1:7])

#remove individual GAD scores
gsd <- subset(gsd, select = -c(1:7))

###SWL Scores

#sum SWL Scores
gsd$Total_SWL_Score <- rowSums(gsd[2:6])

#remove individual SWL Scores
gsd <- subset(gsd, select = -c(2:6))


###SPIN Scores

#sum scores
gsd$Total_Spin_Score <- rowSums(gsd[10:26])

#remove individual SPIN scores
gsd <- subset(gsd, select= -c(10:26))

```

### Modifying GAD, SWL, and SPIN scores for analysis
These assumptions are made following the guidelines for scoring each survey.
```{r}
###GAD - general anxiety disorder

#GAD summed meaning: a total score greater then 10 means there is likelihood
#of an anxiety disorder. for our research we will assume this is true
#and assign a new column "Anxiety_Disorder" with:
#0 as minimal anxiety
#1 as mild anxiety
#2 moderate anxiety
#3 sever anxiety
gsd$Anxiety_Disorder <- 0
gsd$Anxiety_Disorder[gsd$Total_GAD_Score < 5] <- 0
gsd$Anxiety_Disorder[gsd$Total_GAD_Score >= 5 & gsd$Total_GAD_Score < 10] <- 1
gsd$Anxiety_Disorder[gsd$Total_GAD_Score >= 10 & gsd$Total_GAD_Score < 15] <- 2
gsd$Anxiety_Disorder[gsd$Total_GAD_Score >= 15] <- 3



###SWL - satisfaction with life

gsd$Life_Satisfaction <- 0

#assign 1 to score of less than 10. 1 means extremely dissatisfied with life
gsd$Life_Satisfaction[gsd$Total_SWL_Score < 10] <- 1

#assign 2 to score between 10 & 15. 2 means generally dissatisfied
gsd$Life_Satisfaction[gsd$Total_SWL_Score >= 10 & gsd$Total_SWL_Score < 15] <- 2

#assign 3 to score between 15 & 20. 3 means slightly below average satisfaction
gsd$Life_Satisfaction[gsd$Total_SWL_Score >= 15 & gsd$Total_SWL_Score < 20] <- 3

#assign 4 to score between 20 & 25. 4 means average satisfaction
gsd$Life_Satisfaction[gsd$Total_SWL_Score >= 20 & gsd$Total_SWL_Score < 25] <- 4

#assign 5 to score between 25 & 30. 5 means satisfied
gsd$Life_Satisfaction[gsd$Total_SWL_Score >= 25 & gsd$Total_SWL_Score < 30] <- 5

#assign 6 to score greater then 30. 6 means greatly satisfied
gsd$Life_Satisfaction[gsd$Total_SWL_Score >= 30] <- 6



### SPIN - Social Phobia Index

gsd$Social_Phobia <- 0

#assign 0 to score of less than 20. 0 means no social phobia
gsd$Social_Phobia[gsd$Total_Spin_Score <= 20] <- 0

#assign 1 to score between 20 and 30. 1 means mild social phobia
gsd$Social_Phobia[gsd$Total_Spin_Score > 20 & gsd$Total_Spin_Score <= 30] <- 1

#assign 2 to score between 30 and 40. 2 means moderate social phobia
gsd$Social_Phobia[gsd$Total_Spin_Score > 30 & gsd$Total_Spin_Score <= 40] <- 2

#assign 3 to score between 40 and 50. 3 means sever social phobia
gsd$Social_Phobia[gsd$Total_Spin_Score > 40 & gsd$Total_Spin_Score <= 50] <- 3

#assign 4 to score greater than 50. 4 means very severe social phobia
gsd$Social_Phobia[gsd$Total_Spin_Score > 50] <- 4



```

## Analysis:

### General Overview of Survey Responses


#### Create Histogram Plots for GAD, SWL, and SPIN
```{r}
#GAD Histogram
gsd$Low_Anxiety <- as.factor(gsd$Total_GAD_Score < 10)
GAD_hist <- ggplot(gsd, aes(x=Total_GAD_Score, fill = Low_Anxiety)) + geom_histogram() +
  ggtitle("Total GAD Score Frequency")


#SWL Histogram
gsd$High_Satisfaction <- as.factor(gsd$Life_Satisfaction>3)
SWL_hist <- ggplot(gsd, aes(x=Life_Satisfaction, fill=High_Satisfaction)) + geom_histogram() + 
  ggtitle("Life Satisfaction Frequency")



#SPIN Histogram
gsd$Low_Social_Phobia <- as.factor(gsd$Social_Phobia <= 1)
SPIN_hist <-  ggplot(gsd, aes(x=Social_Phobia, fill=Low_Social_Phobia)) + geom_histogram() + 
  ggtitle("Social Phobia Score Frequency")


```

#### Histograms

```{r}
#Histogram Grids
grid.arrange(GAD_hist,SPIN_hist,SWL_hist)
```

The plots above show the frequency distributions of the responses to the 
survey questions regarding GAD - General Anxiety Disorder, SPIN - Social Phobia
Inventory, and SWL - Satisfaction with Life.

**GAD**
We see that scores with less than 10 mean a low likelihood of an anxiety 
disorder shown in blue. The majority of the participants fall into that 
category. 

**SPIN**
A score of 0 or 1 indicate low social phobia, where a high majority of the
participants stand. 

**SWL**
These results show a pretty even distribution of satisfied and unsatisfied
participants with their lives. A score of 3 or less indicates low satisfaction.
Note that the value of the scores in this question are inverted compared to the
previous two. 

### Comparing Hours Gaming and Scores

In this section we will explore the relationship between hours spent gaming per
week and scores.

First, look at the data on hours spent gaming.

```{r}
summary(gsd$Hours)
hours_hist <- hist(gsd$Hours)
mean_hours <- mean(gsd$Hours, na.rm=TRUE)
```
The average of hours spent gaming per week is 21.5. 


**GAD and Hours**

Here, we explore the data by plotting GAD, SPIN, and SWL scores along with hours spent gaming.
```{r}
gsd_hours_GAD <- gsd %>% group_by(Total_GAD_Score) %>% 
  summarise(avg_hours = mean(Hours))
p1 <- ggplot(data= gsd_hours_GAD, aes(x=avg_hours, y=Total_GAD_Score)) + geom_line() +
  geom_smooth(span =1, method=lm) + ggtitle("Total GAD Scores vs Avg Hours Played")

p1a <- ggplot(data= gsd, aes(x=Hours, y=Total_GAD_Score)) + geom_point() +
  geom_smooth(span =1) + ggtitle("Total GAD Scores vs Hours Played")
p1a
p1
```

Here we have 2 graphs, one with the total hours and GAD scores and the other
with averaged hours spent per week by GAD scores. Both show a positive trend,
with the averaged being much more extreme. This tells us two things:
1. There is a positive trend with time spent gaming and general anxiety
2. When participants are grouped by their GAD scores with the average taken
of their hours played, GAD shows a positive relationship.

**SPIN and Hours**
```{r}
gsd_hours_SPIN <- gsd %>% group_by(Social_Phobia) %>% 
  summarise(avg_hours = mean(Hours))
p2 <- ggplot(gsd_hours_SPIN, aes(x=avg_hours, y=Social_Phobia)) + 
  geom_smooth(method=lm) + ggtitle("Social Phobia Scores vs Avg Hours Played")
p2a <- ggplot(gsd, aes(x=Hours, y=Social_Phobia)) + 
  geom_smooth(method=lm) + ggtitle("Social Phobia Scores vs Hours Played")
p2a
p2
```
The social phobia scores show another positive trend with average hours played
per week.

```{r}
gsd_hours_SWL <- gsd %>% group_by(Life_Satisfaction) %>% 
  summarise(avg_hours = mean(Hours))
p3 <- ggplot(gsd_hours_SWL, aes(x=avg_hours, y=Life_Satisfaction)) +
  geom_smooth(method=lm) + ggtitle("Life Satisfaction Scores vs Avg Hours Played")
p3a <- ggplot(gsd, aes(x=Hours, y=Life_Satisfaction)) + 
  geom_smooth(method=lm) + ggtitle("Life Satisfaction Scores vs Hours Played")
p3a
p3
```
The life satisfaction plot shows a negative trend, meaning as the hours and 
average hours per week increases, life satisfaction decreases. 

```{r}
grid.arrange(p1,p2,p3)
```


```{r}
above_avg_hours <- gsd %>% filter(Hours < 22)
below_avg_hours <- gsd %>% filter(Hours >=22)
```

## Modeling
Since we noticed in our analysis so far, people who spend around
the average amount of hours playing per week are generally not very anxious,
social phobic, or have low life satisfaction. We can see changes above the mean
for GAD, however. So in this section we will look at modeling to explore if GAD
scores are significantly changed when looking above and below the mean. 
Additionally, we check SPIN and sWL for anything interesting.

**GAD** Score vs Hours Correlation
```{r}
cor_GAD<-cor.test(gsd$Hours, gsd$Total_GAD_Score, 
                  method =c("pearson", "kendall","spearman"))
cor_GADabove_avg_hours <- cor.test(above_avg_hours$Total_GAD_Score, 
         above_avg_hours$Hours, method =c("pearson", "kendall","spearman"))
cor_GADbelow_avg_hours <- cor.test(below_avg_hours$Total_GAD_Score, 
         below_avg_hours$Hours, method =c("pearson", "kendall","spearman"))
cor_GAD
cor_GADabove_avg_hours
cor_GADbelow_avg_hours
```
The highest coefficient here is for gamers with below average games per week. 
This is interesting because it goes against our hypothesis that spending a lot
of time gaming increases anxiety. Looking at the line graph 
**Total GAD Scores vs AVG Hours Played**, there is a dip at around the average
amount of time spent gaming which would theoretically show a negative
correlation below the average. However, looking at the individual hours played
per participant, there is still a positive relationship. 


**SPIN** Score vs Hours Correlation
```{r}
cor_SPIN<-cor.test(gsd$Hours, gsd$Social_Phobia, 
                  method =c("pearson", "kendall","spearman"))
cor_SPINabove_avg_hours <- cor.test(above_avg_hours$Social_Phobia, 
         above_avg_hours$Hours, method =c("pearson", "kendall","spearman"))
cor_SPINbelow_avg_hours <- cor.test(below_avg_hours$Social_Phobia, 
         below_avg_hours$Hours, method =c("pearson", "kendall","spearman"))
cor_SPIN
cor_SPINabove_avg_hours
cor_SPINbelow_avg_hours

```

Looking at the correlations of the SPIN data, we see something interesting. For
all gamers, the correlation is higher meaning there is a greater positive
relationship with social phobia than participants who spend above average time 
gaming per week. From this, we can gather that those who spend a lot of time
gaming are less likely to experience social phobia. This could be explained by
the fact that these games are online, allowing for socialization. 

**SWL** Score vs Hours Correlation
```{r}
cor_SWL<-cor.test(gsd$Hours, gsd$Life_Satisfaction, 
                  method =c("pearson", "kendall","spearman"))
cor_SWLabove_avg_hours <- cor.test(above_avg_hours$Life_Satisfaction, 
         above_avg_hours$Hours, method =c("pearson", "kendall","spearman"))
cor_SWLbelow_avg_hours <- cor.test(below_avg_hours$Life_Satisfaction, 
         below_avg_hours$Hours, method =c("pearson", "kendall","spearman"))
cor_SWL
cor_SWLabove_avg_hours
cor_SWLbelow_avg_hours
```
Finally, we look at life satisfaction. We find a similar result to SPIN, except
it is a negative relationship. This makes since because a high score is 
associated with high satisfaction. The P-values for these correlations is very
low, so the findings cannot be soundly explained. What we can gather, is that
those who spend above average time gaming are less likely to be dissatisfied
with life. The way we may be able to think about this that people who game
often could get some satisfaction out of it.

##Conclusion

Overall, we can conclude that time spent gaming has an effect on a persons potential for General Anxiety, Social Phobia and Dissatisfaction in life. However, we can't say that gaming is the cause of these issues. A few predictions on why we can't confirm our hypothesis are:

1. Not enough data

Our data set is comprised with a vast majority of League of Legends players. There are over 100 million active LoL players, of which our data contains ~11,000 of, is barely scraping the surface. 

2. General Population

Since we don't have GAD, SPIN, and SWL data for the general population (i.e not all gamers), we don't have anything to compare the gamers' scores to. How do we know how anxious, social phobic, and dissatisfied with life people are in general? It's possible that our data could be a reflection of the general population.
